15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
updates to discussion

---
title: ''
output: 
    #pdf_document: 
    bookdown::word_document2: 
      md_extensions: +raw_html-markdown_in_html_blocks
      pandoc_args: ['--lua-filter', 'read_html.lua']
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: citations.bib
csl: new-phytologist.csl

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first focal area we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This area has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second portion is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third section of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

@@

Consistent with hypotehsis 3, deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5). The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^. In particular, the recruitment rate of *Carya* spp. and *Quercus* spp., late-successional taxa characteristic of this forest's expected climax state, was reduced by > 50% in areas with high deer density. The prevalence of vulnerable canopy species in regions with low deer density showed no significant impact on the total quantity of canopy species recruitment into the understory. However, recruitment of *Fraxinus americana* and *Fagus grandifolia* was higher in areas with high canopy vulnerability, while recruitment of *Ulmus rubra* and *Carya* spp. was lower. Across all regions, the recruitment rate of understory taxa was nearly 10 times that of canopy taxa. Recruitment of *Lindera benzoin* was high across all regions (> 100 stems ha^-1^ yr^-1^), and significantly increased in areas with high deer densities while peaking in areas that also had low canopy vulnerability. *Asimina triloba* recruitment was also lowest within the deer exclosure (`r round(group1_astrRecruit,1)` stems ha^-1^ yr^-1^) but was highest in the region with high deer density and high canopy vulnerability (`r round(group3_astrRecruit,1)` stems ha^-1^ yr^-1^).

@@

Relying on the current understory as a proxy for future canopy presents challenges.
White-tailed deer populations have varied greatly over the last century [@citation].
Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events (Koenig, 2021).
A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse could allow for substantial, rapid regeneration of canopy species.
However, current recruitment levels are below a rate which we would expect in maturing secondary forests.
When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting broader applicability.
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).

>>>>

---
title: '15yrsChange'
output: 
    #pdf_document: 
    bookdown::word_document2: 
      md_extensions: +raw_html-markdown_in_html_blocks
      pandoc_args: ['--lua-filter', 'read_html.lua']
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: citations.bib
csl: new-phytologist.csl

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first focal area we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This area has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second portion is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third section of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

@@

Consistent with hypothesis 3, deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5). The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^. In particular, the recruitment rate of *Carya* spp. and *Quercus* spp., late-successional taxa characteristic of this forest's expected climax state, was reduced by > 50% in areas with high deer density. The prevalence of vulnerable canopy species in regions with low deer density showed no significant impact on the total quantity of canopy species recruitment into the understory. However, recruitment of *Fraxinus americana* and *Fagus grandifolia* was higher in areas with high canopy vulnerability, while recruitment of *Ulmus rubra* and *Carya* spp. was lower. Across all regions, the recruitment rate of understory taxa was nearly 10 times that of canopy taxa. Recruitment of *Lindera benzoin* was high across all regions (> 100 stems ha^-1^ yr^-1^), and significantly increased in areas with high deer densities while peaking in areas that also had low canopy vulnerability. *Asimina triloba* recruitment was also lowest within the deer exclosure (`r round(group1_astrRecruit,1)` stems ha^-1^ yr^-1^) but was highest in the region with high deer density and high canopy vulnerability (`r round(group3_astrRecruit,1)` stems ha^-1^ yr^-1^).

@@

Relying on the current understory as a proxy for future canopy presents challenges.
White-tailed deer populations have varied greatly over the last century [@hanberry_regaining_2020].
Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events [@koenig_brief_2021].
A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse could allow for substantial, rapid regeneration of canopy species.
However, current recruitment levels are below a rate which we would expect in maturing secondary forests.
When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context. EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list). The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022]. Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threats_2020].
